---
title: "Project Network-based Data Analysis"
author: "Annalisa Xamin"
output: 
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
always_allow_html: true
---
```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
# knitr::opts_chunk$set(cache = TRUE)
setwd("~/Documents/Repo_Git/NBDA/Code")
```

---

## Install and load R packages

```{r packages, warning=FALSE, results='hide', message=FALSE}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!require("GEOquery", quietly = TRUE))
    BiocManager::install("GEOquery") 

if (!require("rScudo", quietly = TRUE))
    BiocManager::install("rScudo") 

library(tidyverse)
library(dplyr)
library(GEOquery)
library(rScudo)
library(plotly)
library(sessioninfo)
library(ggplot2)
library(factoextra)
library(useful)

my_colors <- c("#2e005d", "#5c008b", "#8e008b", "#ff8300", "#ff6200", "#d1105a", "#05a8aa")
```

# R session information
Print the R session information for reproducibility.
```{r session info, warning=FALSE}
print('Reproducibility information:') 
Sys.time()
proc.time()
options(width = 120)
session_info()
```

# Data selection

The analysis will use the dataset [GSE20437](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE20437) obtained from GEO.The dataset is generated from Affymetrix HU133A microarrays and contains 42 tissue samples. 

In detail, the data includes:

- 18 reduction mammoplasty (RM) breast epithelium samples, 

- 18 histologically normal (HN) epithelial samples from breast cancer patients (9 ER+ and 9 ER-), and

- 6 histologically normal epithelial samples from prophylactic mastectomy patients. 

Note that sample numbers correspond to individual patient samples.

```{r retrieve data from GEO}
# download the GSE20437 expression data series
gse <- getGEO("GSE20437", destdir= './data/', getGPL = F)
```

```{r}
# getGEO returns a list of expression objects, but...
length(gse) 
# shows us there is only one object in it. 
# We assign it to the same variable.
gse <- gse[[1]]
```

```{r retrieve metadata}
# extract metadata
metadata <- data.frame(gse@phenoData@data)

# save metadata to csv file
#write.csv(metadata, file = 'metadata.csv', row.names = TRUE)
```

```{r plot groups, warning=FALSE}
p <- ggplot(metadata, aes(x=disease.state.ch1, fill=specimen.ch1))+
  geom_bar()+
  scale_fill_manual(values = my_colors[c(1,4,6,7)])
p + labs(x = "Group") 

```


```{r, include=FALSE}
head(metadata)
```


# Exploratory analysis
```{r}
# show what we have:
show(gse)
```

The actual expression data are accessible in the `exprs` section of `gse`, an Expression Set and the generic data class that BioConductor uses for expression data.

```{r inspect exprs}
head(exprs(gse)) 
```

```{r length exprs, include = FALSE}
length(exprs(gse))
```

To conveniently access the data rows and columns present in `exprs(gse)`, this matrix is assigned to its own variable `ex`.

```{r}
# exprs (gse) is a matrix that we can assign to its own variable, to
# conveniently access the data rows and columns
ex <- exprs(gse)
dim(ex) # 42 sample, 22283 genes
```

The dataset contains gene expression data of 22283 genes (rows) from 42 patients (columns).

```{r, include=FALSE}
colnames(ex)
```

```{r, include = FALSE}
rownames(ex)
```

```{r, include = FALSE}
ex[1:5,]
```

```{r, include=FALSE}
metadata$description
# contains case number
```


## Pre-processing

```{r boxplot original data, warning=FALSE}
# Analyze value distributions
boxplot(ex, main = 'Boxplot of the data before normalization',
        xlab = "Samples",
        ylab = "Expression Value",
        varwidth = TRUE
        )
```

The boxplot shows that scaling is necessary. So, in this case, I try to apply a log transformation to the data.

```{r boxplot log transformation, fig.cap="Boxplot of the data after applying a logarithmic transformation"}
ex2<-log(ex)
ex2 <- na.omit(as.matrix(ex2))
#dim(ex2) # 22283    42 same as before
boxplot(ex2, main = 'Boxplot of the data after applying a logarithmic transformation',
        xlab = "Samples",
        ylab = "Expression Value"
        )
```

From the boxplot after the log transformation, I can see that there is some variation in the median of the samples. So, I try to apply a median normalization to the data after the log transformation.

```{r median normalization and boxplot, fig.cap="Boxplot of the data after median normalization"}
# MEDIAN NORMALIZATION
channel.medians=apply(log(ex),2,median)
normalized.log.ex=sweep(log(ex),2,channel.medians,"-")

# boxplot post median normalization on ex
boxplot(normalized.log.ex, 
        main = 'Boxplot of the data after median normalization',
        xlab = "Samples",
        ylab = "Expression Value")
```

## PCA
PCA is a dimensionality reduction technique that allows to condense thousands of dimensions into just two or three. For the dataset's samples, the PCA scores display the coordinates in relation to these additional dimensions. 

```{r pca}
pca <- prcomp(t(normalized.log.ex))

summary(pca)
screeplot(pca)
```

To get the summary of the PCA and the plot showing the variance explained by the first 10 components, it is possible to use the functions commented in the chunks above.

However, using `ggplot2` and `factoextra` packages is possible to get a more concise and informative plot reporting the same information.

```{r variance explained by pca}
pcaVar <- get_eig(pca)
pcaVar <- pcaVar$variance.percent[1:10]
screeDf <- data.frame("Dimensions" = as.factor(seq(1,10)),
                      "Percentages" = pcaVar,
                      "Labels" = paste(round(pcaVar, 2), "%"))

p <- ggplot(data = screeDf, aes(x=Dimensions, y=Percentages))+
  geom_bar(stat = "identity", fill = "#d1105a")+
  geom_text(aes(label=Labels), vjust=-0.5, color="black", size=3.6)+
  ggtitle("Scree Plot")+
  ylab("Percentage of variance explained")+
  scale_x_discrete(labels = as.factor(seq(1,10)))
p
```

The scree plot shows that the first dimensions on the left are the more important because the percentage of variance explained by them is higher. The remaining principal components account for a very small proportion of the variability and are probably unimportant.

Let's try to plot the PCA, looking if we can see a separation between Control and Breast Cancer groups.

```{r PCA for components 1 and 2 control VS Breast cancer without labels}
# draw PCA plot control VS breast cancer
group <- c(rep("cadetblue1",18), rep("red",18), rep("cadetblue1",6) ) 
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$data), cex=0.75)
legend("topleft", col=c("cadetblue1","red"), legend = c("Control", "Breast Cancer"),
    pch = 20, bty='n', cex=.55)
```

```{r PCA for components 1 and 2 control VS Breast cancer with labels, include=FALSE}
# draw PCA plot control VS breast cancer
group <- c(rep("cadetblue1",18), rep("red",18), rep("cadetblue1",6) ) 
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$x), cex=0.5) # add labels to points
legend("topleft", col=c("cadetblue1","red"), legend = c("Control", "Breast Cancer"),
    pch = 20, bty='n', cex=.55)
```
Let's try to add the control subtypes. The vector group used in the PCA plot is based on the data. The samples corresponding to the colors are the following:

-   **Light blue**: reduction mammoplasty (RM) breast epithelium samples

-   **Red**: histologically normal (HN) epithelial samples from breast cancer patient

-   **Purple**: histologically normal breast epithelium (NlEpi) from prophylactic mastectomy patient samples

```{r PCA for components 1 and 2 control subtypes without labels}
# draw PCA plot
group <- c(rep("cadetblue1",18), rep("red",18), rep("purple",6) ) # vector of colors based on the order of my data
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$data), cex=0.75)
legend("topleft", col=c("cadetblue1","red","purple"), legend = c("Reduction Mammoplasty", "Breast Cancer", "Prophylactic Mastectomy"),
    pch = 20, bty='n', cex=.55)
```
```{r PCA for components 1 and 2 control subtypes with labels, include=FALSE}
# draw PCA plot
group <- c(rep("cadetblue1",18), rep("red",18), rep("purple",6) ) # vector of colors based on the order of my data
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$x), cex=0.5) # add labels
legend("topleft", col=c("cadetblue1","red","purple"), legend = c("Reduction Mammoplasty", "Breast Cancer", "Prophylactic Mastectomy"),
    pch = 20, bty='n', cex=.55)
```


Then, I try to see if there is a separation also inside different types of Breast Cancer.
```{r pca with all subtypes without labels}
# draw PCA plot with all subtypes
group <- c(rep(my_colors[7],18), rep(my_colors[4],9), rep(my_colors[1],9), rep(my_colors[6],6) ) # vector of colors based on the order of my data
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$data), cex=0.75)
legend("topleft", col=c(my_colors[7],my_colors[4],my_colors[1],my_colors[6]), legend = c("Reduction Mammoplasty", "ER+ Breast Cancer", "ER- Breast Cancer", "Prophylactic Mastectomy"),
    pch = 20, bty='n', cex=.55)
```

```{r pca with all subtypes with labels, include=FALSE}
# draw PCA plot with all subtypes
group <- c(rep(my_colors[7],18), rep(my_colors[4],9), rep(my_colors[1],9), rep(my_colors[6],6) ) # vector of colors based on the order of my data
plot(pca$x[,1], pca$x[,2], xlab="PCA1", ylab="PCA2", main="PCA for components 1 and 2", type="p", pch=10, col=group)
text(pca$x[,1], pca$x[,2], rownames(pca$x), cex=0.5)
legend("topleft", col=c(my_colors[7],my_colors[4],my_colors[1],my_colors[6]), legend = c("Reduction Mammoplasty", "ER+ Breast Cancer", "ER- Breast Cancer", "Prophylactic Mastectomy"),
    pch = 20, bty='n', cex=.55)
```

### Interactive PCA plot
Let's try to explore an interactive PCA plot.

```{r fig1 interactive PCA plot, message=FALSE}
components<-pca[["x"]]
components<-data.frame(components)
type<-c(rep("RM", 18), rep("HN",18), rep("NlEpi",6))
components<-cbind(components, type )

fig <- plot_ly(components, x=~PC1, y=~PC2, 
               color=type,colors=c('cadetblue1', 'red','purple'), 
               type='scatter',mode='markers')
fig
```

```{r fig2 interactive PCA plot, message=FALSE}
fig2 <- plot_ly(components, x=~PC1, y=~PC2, z=~PC3, 
                color=type, colors=c('cadetblue1', 'red','purple'),
                mode='markers', marker = list(size = 4))
fig2
```


```{r fig3 interactive PCA plot, message=FALSE}
fig3 <- plot_ly(components, x=~PC1, y=~PC3, 
                color=type, colors=c('cadetblue1', 'red','purple'),
                type='scatter',mode='markers')
fig3
```

## UMAP

```{r UMAP 2D, include=FALSE}
umap_data <- umap::umap(t(normalized.log.ex), 
                        n_neighbors = sqrt(dim(t(normalized.log.ex))[1]),#or the square root of the rows 
                        min_dist = 0.1,         
                        metric = "euclidean",    #you can change it
                        n_components = 2) #used the default ones! 

umap_df <- data.frame(umap_data$layout)
group_data <- data.frame(metadata$specimen.ch1)
umap_df <- cbind(umap_df,group_data)
colnames(umap_df) <- c("X1", "X2", "Group")

umap_df$Group<-c(rep("Reduction Mammoplasty", 18), rep("ER+ Breast Cancer",9), rep("ER- Breast Cancer", 9), rep("Prophylactic Mastectomy",6))

library(plotly)

figUmap <- plot_ly(umap_df, 
                 x = ~X1, y = ~X2, color = umap_df$type,
                 colors = c('cadetblue1', 'red','purple'),   
                 mode = 'markers',
                 size=1)

# Display the 2D scatter plot
figUmap
```

```{r, include=FALSE}
umap_data <- umap::umap(t(normalized.log.ex), 
  n_neighbors = 15,            # Number of neighboring data points to consider
  metric = "euclidean",        # Distance metric to use
  n_components = 2,            # Number of components in the reduced space
  min_dist = 0.1,              # Minimum distance between points in the embedding
  spread = 1.0,                # Spread of the low-dimensional representation
  set_op_mix_ratio = 1.0,      # Mix ratio for set operation
  local_connectivity = 1.0,    # Connectivity parameter for local UMAP
  repulsion_strength = 1.0,    # Strength of the repulsion forces
)
umap_df <- data.frame(umap_data$layout)
#group_data <- data.frame(metadata$specimen.ch1)
#umap_df <- cbind(umap_df,group_data)
colnames(umap_df) <- c("Umap_1", "Umap_2") 
                       #,"Group")
rownames(umap_df) <-  rownames(t(normalized.log.ex)) 
print(umap_data)

umap_df$Group<-c(rep("Reduction Mammoplasty", 18), rep("ER+ Breast Cancer",9), rep("ER- Breast Cancer", 9), rep("Prophylactic Mastectomy",6))

```


```{r UMAP 3D, include=FALSE}
umap_data_3D <- umap::umap(t(normalized.log.ex), 
                        n_neighbors = sqrt(dim(t(normalized.log.ex))[1]),
                      #or the square root of the rows, provato anche con 5 fa pena, anche 15 abbastanza pena  
                        min_dist = 0.1,         
                        metric = "euclidean",    #you can change it
                        n_components = 3) #used the default ones! 

umap_df_3D <- data.frame(umap_data_3D$layout) %>% tibble::rownames_to_column('Samples')
umap_df_3D$type<-c(rep("RM", 18), rep("HN",18), rep("NlEpi",6))

figUmap_3D <- plot_ly(umap_df_3D, 
                 x = ~X1, y = ~X2, z=~X3, color = umap_df_3D$type,
                 colors = c('cadetblue1', 'red','purple'),   
                 mode = 'markers',
                 size=1) %>% layout(title = 'Umap 1 control-tumors 3D')

# Display the 2D scatter plot
figUmap_3D
```


# Clustering
## K-means
```{r K-means}
set.seed(1)
k <- 2 # number of clusters

kmeans_result <- kmeans(t(normalized.log.ex),k)
table(kmeans_result$cluster) # tells how many samples were assigned to each cluster
```

```{r plot kmeans, include=FALSE}
plot(kmeans_result, data=t(normalized.log.ex)) + geom_text(aes(label=colnames(ex)),hjust=0,vjust=0)
```

```{r plot kmeans with metadata title}
plot(kmeans_result, data=t(normalized.log.ex)) + geom_text(aes(label=metadata$disease.state.ch1),hjust=0,vjust=0)
```

```{r cluster plot kmeans}
fviz_cluster(kmeans_result, data = t(normalized.log.ex),
             palette = c("#2E9FDF", "#E7B800", "green", "red"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )
```

Let's try increasing the number of clusters.
```{r}
set.seed(1)
k <- 4 # number of clusters

kmeans_result <- kmeans(t(normalized.log.ex),k)
table(kmeans_result$cluster) # tells how many samples were assigned to each cluster
```

```{r plot kmeans with metadata specimen}
plot(kmeans_result, data=t(normalized.log.ex)) + geom_text(aes(label=metadata$specimen.ch1),hjust=0,vjust=0)
```

```{r cluster plot kmeans}
fviz_cluster(kmeans_result, data = t(normalized.log.ex),
             palette = c("#2E9FDF", "#E7B800", "green", "red"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )
```


